Open archaeology, open source?

### 2024-02-03 analysis/_01-introduction.qmd
General updates to the text, and also replaced the main data file.
Text modifications throughout. Replaced the data file according to the new release (https://github.com/zackbatist/open-archaeo/releases/tag/2024-02-03). Revised the abstract. Verified and corrected the bibliographic references.

However, academic open source has a complicated relationship with open source as practiced by professional software developers, which has its own distinct history and is framed by different objectives, challenges, and value regimes. Despite this, the open science movement, within which open archaeology emerged, draws direct inspiration from open source. For instance, the Open Knowledge Foundation [-@openknowledgefoundation2015] publishes a widely accepted definition of "open" in the context of scholarly communication that explicitly refers to the definition of "open source" published by the Open Source Initiative [-@opensourceinitiative2007], an authoritative open source advocacy group. The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plain text, line-resolution version control, emphasis on formal licencing, the general hacker aesthetic). However, academic work, including the development of academic software, differs significantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, openSSL, and the Firefox web browser. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional warrants and interests. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers [@ratto2007; @kelty2008 chapter 9].

@@

Despite hopeful aspirations espoused by open archaeology advocates [see @kansa2014; @kintigh2015], these outcomes are not a given. We believe<!-- Bit weak; can we cite some previous work that indicates this? --JR---> that these outcomes only actually arise in contexts where participants adhere to and are motivated by warrants, professional norms, and governance strategies that encourage these results. However, practical circumstances and systemic value regimes that frame what it means to work as an archaeologist presently inhibit the potential for radical transformation, even among open science's most ardent supporters.

<!-- Might be getting a bit ahead of ourselves here. --JR -->That being said, archaeologists are prolific software developers. This is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources that indexes 501 records, as of the time of this article's publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

>>>>

However, academic open source has a complicated relationship with open source as practiced by professional software developers, which has its own distinct history and is framed by different objectives, challenges, and value regimes. Despite this, the open science movement, within which open archaeology emerged, draws direct inspiration from open source. For instance, the Open Knowledge Foundation [-@openknowledgefoundation2015] publishes a widely accepted definition of "open" in the context of scholarly communication that explicitly refers to the definition of "open source" published by the Open Source Initiative [-@opensourceinitiative2007], an authoritative open source advocacy group. The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plain text, line-resolution version control, emphasis on formal licencing, the general hacker aesthetic). However, academic work, including the development of academic software, differs significantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, openSSL and the Firefox web browser. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional warrants and interests. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers [@ratto2007; @kelty2008 chapter 9].

@@

Despite hopeful aspirations espoused by open archaeology advocates [see @kansa2014; @kintigh2015], these outcomes are not a given. As per @nguyen2022, @pownall2023 and @leonelli2023 we believe that these outcomes only actually arise in contexts where participants adhere to and are motivated by warrants, professional norms, and governance strategies that encourage these results. However, practical circumstances and systemic value regimes that frame what it means to work as an archaeologist presently inhibit the potential for radical transformation, even among open science's most ardent supporters.

That being said, archaeologists are prolific software developers. This is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources that indexes 495 records, as of the time of this article's publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

### 2024-02-03 analysis/_03-data_methodology.qmd
General updates to the text, and also replaced the main data file.
Text modifications throughout. Replaced the data file according to the new release (https://github.com/zackbatist/open-archaeo/releases/tag/2024-02-03). Revised the abstract. Verified and corrected the bibliographic references.

```{r data-oarch}
# open-archaeo version 2024-02-02
# https://doi.org/10.5281/zenodo.10611198
oarch <- read_csv(here("analysis/data/raw_data/open-archaeo_2024-02-02.csv"))

>>>>

```{r data-oarch}
# open-archaeo version 2024-02-03
# https://doi.org/10.5281/zenodo.10613694
oarch <- read_csv(here("analysis/data/raw_data/open-archaeo_2024-02-03.csv"))

### 2024-02-03 analysis/_04-open_archaeology.qmd
General updates to the text, and also replaced the main data file.
Text modifications throughout. Replaced the data file according to the new release (https://github.com/zackbatist/open-archaeo/releases/tag/2024-02-03). Revised the abstract. Verified and corrected the bibliographic references.

Our analysis of tags highlights aspects of archaeological work that software developers are inclined to contribute to [@tbl-tags].
The most common tags unsurprisingly deal with work that naturally benefits from advanced information processing afforded by computers, such as statistical analysis, sample calibration, geographical analysis, data management, and chronological modelling.
Educational resources and practical guides are also well represented due to the web's usefulness as a medium for sharing and communication.

@@

Roughly half of open-archaeo repositories are accompanied by an explicit license [@tbl-repo-licenses].
Two common free software licenses account for the majority of these: the GNU General Public License (GPL, `r percent(p_license_gpl)`) and the MIT License (`r percent(p_license_mit)`).
These differ primarily in the restrictions they place on reuse: the MIT License aims to be maximally permissive, while the GPL is a 'copyleft' license that specifies that all derivative works must be distributed under similar terms (in other words, it prohibits the use of FOSS software within non-FOSS software, [@dusollier2007]).
Archaeologists' preference for the more restrictive of these two licenses is the reverse of the general trend seen in open source projects on GitHub [@balter2015].
Creative Commons licenses are a distant third place (`r percent(p_license_cc)` of repositories), in contrast to their widespread use for other forms of scholarly output [@kim2007].
Many repositories do not specify a license; given a documented misconception among academics that GitHub can serve as a sustainable and long-term code and data hosting platform [@milliken2021; @escamilla2022; @escamilla2023], it is possible that many maintainers whose work is included in open-archaeo similarly assumed that making their work available, without explicitly stating permissible use, is enough to allow unrestricted access to the repository's contents.
However, we can not verify this potential explanation given the methods we currently employ, and more discursive qualitative research is needed to explore the rationales behind such decisions.

>>>>

Our analysis of tags highlights aspects of archaeological work that software developers are inclined to contribute to (@tbl-tags).
The most common tags unsurprisingly deal with work that naturally benefits from advanced information processing afforded by computers, such as statistical analysis, sample calibration, geographical analysis, data management, and chronological modelling.
Educational resources and practical guides are also well represented due to the web's usefulness as a medium for sharing and communication.

@@

Roughly half of open-archaeo repositories are accompanied by an explicit license (@tbl-repo-licenses).
Two common free software licenses account for the majority of these: the GNU General Public License (GPL, `r percent(p_license_gpl)`) and the MIT License (`r percent(p_license_mit)`).
These differ primarily in the restrictions they place on reuse: the MIT License aims to be maximally permissive, while the GPL is a 'copyleft' license that specifies that all derivative works must be distributed under similar terms (in other words, it prohibits the use of open source software within non-open software, [@dusollier2007]).
Archaeologists' preference for the more restrictive of these two licenses is the reverse of the general trend seen in open source projects on GitHub [@balter2015].
Creative Commons licenses are a distant third place (`r percent(p_license_cc)` of repositories), in contrast to their widespread use for other forms of scholarly output [@kim2007].
Many repositories do not specify a license; given a documented misconception among academics that GitHub can serve as a sustainable and long-term code and data hosting platform [@milliken2021; @escamilla2022; @escamilla2023], it is possible that many maintainers whose work is included in open-archaeo similarly assumed that making their work available, without explicitly stating permissible use, is enough to allow unrestricted access to the repository's contents.
However, we can not verify this potential explanation given the methods we currently employ, and more discursive qualitative research is needed to explore the rationales behind such decisions.

### 2024-02-03 analysis/_06-network.qmd
General updates to the text, and also replaced the main data file.
Text modifications throughout. Replaced the data file according to the new release (https://github.com/zackbatist/open-archaeo/releases/tag/2024-02-03). Revised the abstract. Verified and corrected the bibliographic references.

<!-- Predictors of repo centrality -->
<!-- TODO: Needs an update. These repos are no longer in the top 10, and it's probably best not to be so specific given how volatile that is. -->
The repositories most central to the network as a whole [@tbl-repo-betweenness] include several lists and directories, including open-archaeo itself.
Four of the top ten most central repositories (ISAAKiel/mortAAR, ISAAKiel/shapAAR, ISAAKiel/quantAAR, ropensci/c14bazAAR) are associated with the ISAA-Kiel group, emphasising the importance of existing 'real world' collaborations in structuring the network.
Three (ropensci/c14bazAAR, ahb108/rcarbon, Chronomodel/chronomodel) are R packages relating to radiocarbon data and modelling, probably reflecting the fact that this has long been a prominent focus of statistical software development in archaeology.
Repository centrality is predicted by the total number of commits it has received but, somewhat surprisingly, it is younger repositories rather than older ones that tend to be more central.
Tentatively, we interpret this as an indication that the network has become more connected over time, but we leave a fuller analysis of temporal trends in collaborative activity to future work.

@@

The highest mean centrality is seen in repositories that contain lists/datasets, standalone software, or packages/libraries, and in repositories based on Python, R or QGIS [@fig-repo-betweenness-category-platform].
Interestingly, these trends depart from the observed popularity of different categories/platform in the open-archaeo dataset as a whole (see @sec-open-archaeology): standalone software is more central than packages/libraries, even though there are more of the latter by a significant margin.
This may be due to the fact that many packages are developed to support specific practices or use-cases (often inspired by personal need), or are designed to run relatively stable statistical functions that need not change over time.
These are therefore relatively stable and require little additional input after release.
On the other hand, as discussed in @sec-open-archaeology, standalone software tend to integrate multiple system components and may evolve over time to add new features or support new workflows.
Moreover, standalone software are generally rooted in longer-term and community-held objectives, and their development may therefore be backed by insitutions with funding and resources to support developers.

@@

<!-- Predictors of user centrality -->
Centrality to the user–user graph is weakly predicted by a user's overall rate of activity, as measured by their total number of contributions [@fig-user-betweenness-contribs]. 
We did not collect demographic data on users that appear in our dataset, but based on our own knowledge of the community we can observe that those highly central to the network tend to be employed in (junior) academic positions, or in a few cases in cultural heritage authorities, rather than specifically as research software engineers.
Such positions tend to not to actively reward or encourage software development, at least not on a par with more traditional academic outputs [@baxter2012], and are increasingly precarious [@cornelius-bell2021].
This obviously poses a serious risk to the sustainability and growth of open source software in archaeology: if the people who occupy central positions in the network cease to be active, then it is likely that the overall network would fragment.
Assessing and mitigating this risk should be a high priority for future research in this area.

>>>>

<!-- Predictors of repo centrality -->
The repositories most central to the network as a whole (@tbl-repo-betweenness) include three lists and directories, including open-archaeo itself.
Three relate to making large data repositories accessible for analysis, and one is a very well-supported field recording application.
Community input is therefore centred on infrastructural projects, including those which index and publicize available tools and resources.
Moreover, three relate to radiocarbon data modelling and two relate to palaeoenvironment reconstruction, which reflects the fact that these have long been prominent foci of statistical software development in archaeology.

Repository centrality is predicted by the total number of commits it has received but, somewhat surprisingly, younger repositories rather than older ones tend to be more central (@fig-repo-betweenness-age-length).
Tentatively, we interpret this as an indication that the network has become more connected over time, but we leave a fuller analysis of temporal trends in collaborative activity to future work.

@@

When comparing across categories and platforms, the highest mean centrality is seen in repositories that contain lists and datasets, standalone software, or packages and libraries, and in repositories based on Python, R or QGIS (@fig-repo-betweenness-category-platform).
Interestingly, these trends depart from the observed popularity of different categories and platforms in the open-archaeo dataset as a whole (see @sec-open-archaeology): standalone software is more central than packages/libraries, even though there are more of the latter by a significant margin.
This may be due to the fact that many packages are developed to support specific practices or use-cases (often inspired by personal need), or are designed to run relatively stable statistical functions that need not change over time.
These are therefore relatively stable and require little additional input after release.
On the other hand, as discussed in @sec-open-archaeology, standalone software tend to integrate multiple system components and may evolve over time to add new features or support new workflows.
Moreover, standalone software are generally rooted in longer-term and community-held objectives, and their development may therefore be backed by insitutions with funding and resources to support developers.

@@

<!-- Predictors of user centrality -->
Centrality to the user–user graph is weakly predicted by a user's overall rate of activity, as measured by their total number of contributions (@fig-user-betweenness-contribs). 
We did not collect demographic data on users that appear in our dataset, but based on our own knowledge of the community we can observe that those highly central to the network tend to be employed in (junior) academic positions, or in a few cases in cultural heritage authorities, rather than specifically as research software engineers.
Such positions tend to not to actively reward or encourage software development, at least not on a par with more traditional academic outputs [@baxter2012], and are increasingly precarious [@cornelius-bell2021].
This obviously poses a serious risk to the sustainability and growth of open source software in archaeology: if the people who occupy central positions in the network cease to be active, then it is likely that the overall network would fragment.
Assessing and mitigating this risk should be a high priority for future research in this area.

### 2024-02-03 analysis/_07-conclusion.qmd
General updates to the text, and also replaced the main data file.
Text modifications throughout. Replaced the data file according to the new release (https://github.com/zackbatist/open-archaeo/releases/tag/2024-02-03). Revised the abstract. Verified and corrected the bibliographic references.

Our network analysis similarly draws attention to the real-world collaborative ties that underpin archaeological open source software development.
We identify a core cluster representing a series of collaboratibe ties among members of a archaeological software developer community of practice.
This core exhibits complexity that correspond with social patterns.
Indeed, we demonstrate that real-world social connections and institutional support structures are strong predictors of centrality, with sub-clusters appearing within the central core mapping onto established partnerships.
This indicates that archaeological open source is firmly embedded within existing power structures that permeate academic life, both online and offline.

There is thus little evidence to support the notion that archaeologists benefit from the positive outcomes that are commonly argued to be the natural results of open source development models -- namely, greater degrees of transparency, extensibility and participatory action.
While opening the source code may facilitate these positive outcomes as necessary preconditional factors, we argue that this only amounts to establishing the _potential_ for people to put these values into practice. 
Moreover, we argue that the objectives and circumstances that frame archaeological practice significantly influence how far archaeologists (and academics in general) are willing to push for these values, and limit the ability for archaeologists to do open source in ways that resemble more mainstream open source projects.
For instance, successful open source projects like the Linux kernel or the Firefox web browser are driven by collective and popular interest in ensuring that code remains functional, and the code base is therefore constantly in flux and bears an accumulating list of contributing members.
This significantly differs from the organizational principles that govern archaeological work, namely that a project's leaders runs the whole show from the top down and operationalize all other actors as instruments of their central directives. 
Moreover, archaeological projects ultimately seek to produce stable textual outcomes bearing clear delineation of authorship and that require no upkeep whatsoever. 
Sustaining an open source project is simply not compatible with the factors that currently drive the momentum behind archaeological work.

>>>>

Our network analysis similarly draws attention to the real-world collaborative ties that underpin archaeological open source software development.
We identify a core cluster representing a series of collaborative ties among members of an archaeological software engineering community of practice.
This core exhibits complexity that corresponds with social patterns, such as the presence of various sub-clusters representing interconnected interest or affinity groups.
Indeed, we demonstrate that real-world social connections and institutional support structures are strong predictors of centrality, since these sub-clusters are representative of established professional partnerships.
This indicates that archaeological open source is firmly embedded within existing power structures that permeate academic life, both online and offline.

There is thus little evidence to support the notion that archaeologists benefit from the positive outcomes that are commonly argued to be the natural results of open source development models -- namely, greater degrees of transparency, extensibility and participatory action.
While opening the source code may facilitate these positive outcomes as necessary preconditional factors, we argue that this only amounts to establishing the _potential_ for people to put these values into practice. 
Moreover, we argue that the objectives and circumstances that frame archaeological practice significantly influence how far archaeologists (and academics in general) are willing to push for these values, and limit the ability for archaeologists to do open source in ways that resemble more mainstream open source projects.
For instance, successful open source projects like the Linux kernel, openSSL the Firefox web browser are driven by collective and popular interest in ensuring that code remains functional, and the code base is therefore constantly in flux and bears an accumulating list of contributing members.
This significantly differs from the organizational principles that govern archaeological work, namely that a project's leaders runs the whole show from the top down and operationalize all other actors as instruments of their central directives. 
Moreover, archaeological projects ultimately seek to produce stable textual outcomes bearing clear delineation of authorship and that require no upkeep whatsoever. 
Sustaining an open source project is simply not compatible with the factors that currently drive the momentum behind archaeological work.

### 2024-02-03 analysis/openarchaeo_collaboration.qmd
General updates to the text, and also replaced the main data file.
Text modifications throughout. Replaced the data file according to the new release (https://github.com/zackbatist/open-archaeo/releases/tag/2024-02-03). Revised the abstract. Verified and corrected the bibliographic references.

---
title: Open archaeology, open source?
subtitle: Collaborative practices in an emerging community of archaeological software engineers
author:
  - name: Zack Batist
    url: https://zackbatist.info
    affiliation: Faculty of Information, University of Toronto
    affiliation-url: https://ischool.utoronto.ca/
    orcid: 0000-0003-0435-508X 
  - name: Joe Roe
    url: https://joeroe.io
    affiliation: Institute of Archaeological Sciences, University of Bern
    affiliation-url: https://www.iaw.unibe.ch/
    orcid: 0000-0002-1011-1244
date: "`r Sys.Date()`"
abstract: |
  Surveying the first quarter-century of computer applications in archaeology, Scollar [-@scollar1999] lamented that the field relied almost exclusively on "hand-me-down" tools repurposed from other disciplines.  25 years later, this is no longer the case: computational archaeologists often find themselves practicing the dual roles of data analyst and research software engineer [@baxter2012], developing and applying new tools that are tailored specifically to archaeological problems and archaeological methods. Though this trend can be traced to the very earliest days of the field [@cowgill1967], its most recent manifestation is distinguished by its apparently embrace of practices from free and open source software (FOSS). Most prominently, since around 2015, there has been a rapid uptake of workflow tools designed for open source development communities, such as the version control system `git` and associated online source code management platforms (e.g. GitHub, GitLab). These tools facilitate collaboration among developers and users of FOSS software using patterns that can diverge quite radically from conventional scholarly norms [@tennant2020].

  In this paper, we investigate modes of collaboration in this emerging community of practice using open-archaeo [open-archaeo.info](open-archaeo.info), a curated list of 500+ active open projects in archaeology, augmented where available with data on their activity on GitHub. We apply network analysis to map the networks of collaboration that have emerged among archaeologists using GitHub, and conduct a quantitative exploratory analysis to characterize the nature and intensity of these collaborations. Our results show an uneven adoption of open source collaborative practices beyond the basic use of `git` as a version control system and GitHub to host source code. A majority of repositories have 1–3 contributors, with only a few projects distinguished by an active and diverse developer base. More community-oriented features, such as GitHub's 'issues' and 'comments', are used in only a minority of repositories. Preliminary findings from ongoing work also shows how collaborative behaviours vary across different subcommunities. By analyzing interactions among archaeologists using the GitHub platform, including the ways they contribute to collective code bases, we demonstrate how archaeological software engineering is beginning to foster new kinds of collaborative commitments that blend open source protocols, DIY work ethic, conventional forms of scholarly communication, and cultural practices unique to archaeology.
bibliography: references.bib
format:
  html:
    page-layout: article
    toc: true
    number-sections: true
    embed-resources: true
---

>>>>

---
title: Open archaeology, open source?
subtitle: Collaborative practices in an emerging community of archaeological software engineers
author:
  - name: Zachary Batist
    url: https://zackbatist.info
    affiliation: Independent Scholar
    orcid: 0000-0003-0435-508X 
  - name: Joe Roe
    url: https://joeroe.io
    affiliation: Institute of Archaeological Sciences, University of Bern
    affiliation-url: https://www.iaw.unibe.ch/
    orcid: 0000-0002-1011-1244
date: "`r Sys.Date()`"
abstract: |
  Surveying the first quarter-century of computer applications in archaeology, Scollar [-@scollar1999] lamented that the field relied almost exclusively on "hand-me-down" tools repurposed from other disciplines.  25 years later, this is no longer the case: computational archaeologists often find themselves practicing the dual roles of data analyst and research software engineer [@baxter2012; @schmidt2020], developing and applying new tools that are tailored specifically to archaeological problems and archaeological methods. Though this trend can be traced to the very earliest days of the field [@cowgill1967], its most recent manifestation is distinguished by its apparent embrace of practices from free and open source software . Most prominently, since around 2015, there has been a rapid uptake of workflow tools designed for open source development communities, such as the version control system git and associated online source code management platforms (e.g. GitHub, GitLab). These tools facilitate collaboration among developers and users of open source software using patterns that can diverge quite radically from conventional scholarly norms [@tennant2020].

  In this paper, we investigate modes of collaboration in this emerging community of practice using open-archaeo [open-archaeo.info](open-archaeo.info), a curated list of archaeological software, augmented where available with data on their activity on GitHub.
  We apply network analysis to map the networks of collaboration that have emerged among archaeologists using GitHub, and conduct a quantitative exploratory analysis to characterize the nature and intensity of these collaborations.
  Our results show uneven adoption of open source collaborative practices beyond the basic use of git as a version control system and GitHub to host source code.
  For instance, a majority of repositories have 1–3 contributors, with only a few projects distinguished by an active and diverse developer base.
  Furthermore, community-oriented features, such as GitHub's 'issues' and 'comments', are used in only a minority of repositories.
  Our network analysis similarly demonstrates that archaeological open source is firmly embedded within existing power structures that permeate academic life, and that apparent connections are representative of established professional partnerships.
  As such, we find little evidence to support the notion that archaeologists benefit from the potential positive outcomes that are commonly argued to be the natural results of open source development models – namely, greater degrees of transparency, extensibility and participatory action.
bibliography: references.bib
format:
  html:
    page-layout: article
    toc: true
    number-sections: true
    embed-resources: true
---
